---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---


## Loading and preprocessing the data

Note: preprocessing not persofrmed here. Further improvements: for time intervals, colon can be introduced, i.e. "0" => "00:00", "5" => "00:05", 835 => "8:35", "2355" => "23:55" and so on.
```{r setoptions,echo=TRUE}
suppressPackageStartupMessages(library(sqldf));
suppressPackageStartupMessages(library(tcltk));
library(chron);
options(scipen = 6, digits = 2)

if (!exists("rawData")){
    rawData <- read.csv("activity.csv");
}
```


## What is mean total number of steps taken per day?
```{r}
r1 <- sqldf("select date, sum(steps) as sum_steps 
            from rawData 
            where steps IS NOT NULL 
            group by date");
r1mean <- mean(r1$sum_steps);
r1median <- median(r1$sum_steps);
hist(r1$sum_steps, xlab = "number of steps", main = "total number of steps per day");
```
NA values were ignored. So, mean and median of the total number of steps taken per day are `r r1mean` and `r r1median` accordingly.


## What is the average daily activity pattern?
```{r}
dailyPattern <- sqldf("select interval, avg(steps) as avg_steps 
            from rawData 
            where steps IS NOT NULL 
            group by interval
            order by interval asc");
maxStepsIndex <- which.max(dailyPattern$avg_steps);
maxStepsInterval <- dailyPattern$interval[maxStepsIndex];
plot(dailyPattern$interval, dailyPattern$avg_steps, ylab = "number of steps", xlab = "time", main ="average number of steps per 5-minute interval within a day", type = "l");
```
Most "active" interval within a day (i.e. contains the maximum number of steps) is `r maxStepsInterval`


## Imputing missing values
NA values are imputed using average daily pattern(since mean/media for the day is too rough)
```{r}
na_count <- sum(is.na(rawData$steps));
alteredData <- read.csv("activity.csv");
rowLimit <- nrow(alteredData);
for (i in 1:rowLimit){
    if (is.na(alteredData$steps[i])){
        alteredData$steps[i] <- dailyPattern$avg_steps[ alteredData$interval[i] == dailyPattern$interval ];
    }
}

alteredDailySteps <- sqldf("select date, sum(steps) as sum_steps 
            from alteredData 
            group by date");

alteredMean <- mean(alteredDailySteps$sum_steps);
alteredMedian <- median(alteredDailySteps$sum_steps);
hist(alteredDailySteps$sum_steps, xlab = "number of steps", main = "total number of steps per day");
```
Now mean and median of the total number of steps taken per day are `r alteredMean` and `r alteredMedian` accordingly. Values are almost the same (compared to step 2, "What is the average daily activity pattern") since NA values were imputed using average daily pattern.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
alteredData$isWeekday <- factor(ifelse(weekdays(as.POSIXct(strftime(alteredData$date, format = "%Y-%m-%e"))) %in% c("Saturday", "Sunday"), "WeekEnd", "WeekDay"));


alteredDailyPattern <- sqldf("select interval, isWeekday, avg(steps) as avg_steps
            from alteredData 
            where steps IS NOT NULL 
            group by interval, isWeekday
            order by interval asc");

library(ggplot2);
ggplot(alteredDailyPattern, 
       aes(interval, avg_steps) ) + 
    geom_line() + 
    facet_wrap(~ isWeekday, ncol = 1) + 
    xlab("time") + 
    ylab("number of steps") + 
    ggtitle("Differences in activity patterns between weekdays and weekends")
```

Interesting to see, for WeekDays there is a splash at "5:00". WeekEnds have no such splash.
